---
title: "Isopair: Characterizing structural differences between isoform pairs"
author: "Pete Castaldi"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Isopair: Characterizing structural differences between isoform pairs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
eval_plot <- requireNamespace("ggplot2", quietly = TRUE)
eval_meta <- requireNamespace("metafor", quietly = TRUE)
```

# Introduction

How do two isoforms of the same gene differ structurally? **Isopair** answers
this question by detecting, characterizing, and comparing structural differences
between isoform pairs. The package identifies 12 types of splicing events
(alternative TSS/TES, alternative 5'/3' splice sites, skipped exons, intron
retention, and partial intron retention variants), reconstructs reference
isoforms from detected events, and builds splicing choice profiles for
downstream analysis.

Use cases include NMD target characterization, alternative splicing surveys,
and cross-condition comparisons of isoform structure. The workflow proceeds
from event detection through profiling to single-set characterization and
cross-set comparison.

# Setup

```{r library}
library(Isopair)
```

Isopair ships with example data from 5 genes (SRSF3, SRSF1, BCL2L1, TP63,
FAM13A) extracted from GENCODE v49. Three pre-computed objects are available:

```{r load-data}
data(example_structures)  # parsed isoform structures
data(example_pairs)       # expression-derived pairs
data(example_profiles)    # splicing choice profiles
```

The `example_structures` tibble contains `r nrow(example_structures)` isoforms.
The `example_profiles` tibble contains `r nrow(example_profiles)` pair profiles
with `r sum(example_profiles$n_events)` total events.

# Core Workflow

This section walks through the full pipeline from a GTF file to verified
profiles.

## Parse isoform structures

```{r parse}
gtf_path <- system.file("extdata", "example_small.gtf", package = "Isopair")
structures <- parseIsoformStructures(gtf_path, verbose = FALSE)
head(structures[, c("isoform_id", "gene_id", "n_exons")])
```

## Build union exons

Union exons are atomic, non-overlapping segments that serve as a common
coordinate system across all isoforms of a gene. Build them from **all
isoforms** you plan to analyze:

```{r union-exons}
ue <- buildUnionExons(structures, verbose = FALSE)
head(ue$union_exons)
```

## Generate pairs

Load expression data and create isoform pairs. The expression CSV must be
converted to a numeric matrix with isoform IDs as rownames:
```{r pairs}
expr_file <- system.file("extdata", "example_expression.csv",
  package = "Isopair")
expr_df <- read.csv(expr_file)
expr_mat <- as.matrix(expr_df[, -1])
rownames(expr_mat) <- expr_df$isoform_id

gene_map <- unique(as.data.frame(structures[, c("isoform_id", "gene_id")]))
pairs <- generatePairsExpression(expr_mat, gene_map,
  colnames(expr_mat), method = "top_two")
pairs
```

## Build profiles

Profiles detect events for each pair, reconstruct the reference, and verify
the reconstruction:

```{r profiles}
profiles <- buildProfiles(pairs, structures,
  ue$union_exons, ue$isoform_union_mapping, verbose = FALSE)
```

## Verify reconstruction

Every profile includes a reconstruction status. Check that all pairs pass:

```{r verify}
table(profiles$reconstruction_status)
```

## Summarize

```{r summarize}
summarizeProfiles(profiles)
```

# Understanding Events

Isopair detects 12 event types organized in 4 categories:

| Category | Event Types |
|---|---|
| Terminal | Alt_TSS, Alt_TES |
| Boundary | A5SS, A3SS, Partial_IR_5, Partial_IR_3 |
| Exon-level | SE, Missing_Internal |
| Intron retention | IR, IR_diff_5, IR_diff_3, IR_diff_5_3 |

**Direction semantics:** Events are described from the comparator's
perspective. **LOSS** means the comparator lost sequence relative to the
reference (the reference has more). **GAIN** means the comparator gained
sequence (the reference has less).

## Examining detailed events

Each profile row contains a nested `detailed_events` tibble. Access it with
base R:

```{r detailed-events}
profiles$detailed_events[[1]]
```

## Visualization

```{r plot-pair, eval=eval_plot, fig.width=8, fig.height=4}
# Get exon data for the first pair
ref_id <- profiles$reference_isoform_id[1]
comp_id <- profiles$comparator_isoform_id[1]
ref_struct <- structures[structures$isoform_id == ref_id, ]
comp_struct <- structures[structures$isoform_id == comp_id, ]

ref_exons <- data.frame(
  exon_start = ref_struct$exon_starts[[1]],
  exon_end = ref_struct$exon_ends[[1]]
)
comp_exons <- data.frame(
  exon_start = comp_struct$exon_starts[[1]],
  exon_end = comp_struct$exon_ends[[1]]
)

events <- profiles$detailed_events[[1]]
plotIsoformPair(ref_exons, comp_exons, events,
  profiles$gene_id[1], ref_id, comp_id,
  ref_struct$strand)
```

# Generating Pairs

Isopair provides three methods for creating isoform pairs, each suited to
different experimental designs.

## Expression baseline

Pair the two most highly expressed isoforms per gene:

```{r pairs-expr}
pairs_expr <- generatePairsExpression(expr_mat, gene_map,
  colnames(expr_mat), method = "top_two")
nrow(pairs_expr)
```

## Differential expression

First identify dominant isoforms, then pair each significant DE isoform
against its gene's dominant:

```{r pairs-de}
de_file <- system.file("extdata", "example_de.csv", package = "Isopair")
de <- read.csv(de_file)
dom <- identifyDominantIsoforms(expr_mat, gene_map, colnames(expr_mat))
pairs_de <- generatePairsDE(de, "isoform_id", "gene_id", "logFC",
  "adj_p_val", dom)
nrow(pairs_de)
```

## Differential usage

Pair isoforms based on differential usage results:

```{r pairs-du}
du_file <- system.file("extdata", "example_du.csv", package = "Isopair")
du <- read.csv(du_file)
pairs_du <- generatePairsDU(du, "isoform_id", "gene_id", "dPSI",
  "adj_p_val", method = "extreme")
nrow(pairs_du)
```

# Characterizing a Pair Set

This section demonstrates single-set characterization using the pre-computed
`example_profiles`.

## Event quantification

```{r summary}
summarizeProfiles(example_profiles)
```

## Co-occurrence

Test whether event types co-occur more or less often than expected:

```{r cooccurrence}
cooc <- testCooccurrence(example_profiles)
cooc[, c("event_a", "event_b", "odds_ratio", "p_value")]
```

With only `r nrow(example_profiles)` profiles, most tests lack power. In real
analyses with hundreds of pairs, significant co-occurrence patterns emerge.

## Spatial analysis

Test whether events are uniformly distributed along the gene span. We lower
`min_events` for illustration; use the default (30) for real analyses:

```{r positional-bias}
bias <- testPositionalBias(example_profiles, example_structures,
  min_events = 1L)
bias[, c("event_type", "n_events", "median_position", "bias_direction")]
```

### Topology classification

Classify A5SS/A3SS event pairs into Face-to-Face, Back-to-Back, or Distal
topologies:

```{r topology}
topo <- classifyTopology(example_profiles, example_structures, test = FALSE)
topo$summary
```

### Event proximity

Test whether internal events cluster closer than expected by chance:

```{r proximity}
prox <- testProximity(example_profiles, example_structures, n_perm = 50L)
prox[c("z_score", "p_value", "n_profiles_used")]
```

## Regional impact

Map events to genomic regions (5'UTR, CDS, 3'UTR) and test for enrichment:

```{r regional}
cds <- extractCdsAnnotations(gtf_path, structures$isoform_id, verbose = FALSE)
annotated <- annotateRegionTypes(ue$isoform_union_mapping, cds)
er <- mapEventsToRegions(example_profiles, annotated, cds)
head(er[, c("event_type", "region_type")])
```

```{r enrichment}
enrich <- testRegionalEnrichment(er, annotated, example_profiles)
enrich[, c("event_type", "region_type", "observed_prop", "expected_prop",
  "direction_enrich")]
```

## PTC analysis

Compute premature termination codon status and test for association between
pair members:

```{r ptc}
ptc <- computePtcStatus(structures, cds)
head(ptc[, c("isoform_id", "has_ptc", "ptc_distance")])
```

```{r ptc-assoc}
assoc <- testPtcAssociation(ptc, example_pairs)
assoc$ref_ptc_rate
assoc$comp_ptc_rate
```

# Comparing Two Pair Sets

The `comparePairSets()` function runs 10 statistical tests comparing two
profile sets. Each test section is self-documenting: the result includes the
test description, method, p-value, and the exact data that entered the test.

```{r compare-build}
# Build profiles from two different pair methods
profiles_expr <- buildProfiles(pairs_expr, structures,
  ue$union_exons, ue$isoform_union_mapping, verbose = FALSE)

# Use DE pairs (if any exist)
if (nrow(pairs_de) > 0) {
  # Filter to isoforms present in structures
  valid <- pairs_de$reference_isoform_id %in% structures$isoform_id &
    pairs_de$comparator_isoform_id %in% structures$isoform_id
  pairs_de_valid <- pairs_de[valid, ]
  if (nrow(pairs_de_valid) > 0) {
    profiles_de <- buildProfiles(pairs_de_valid, structures,
      ue$union_exons, ue$isoform_union_mapping, verbose = FALSE)
  } else {
    profiles_de <- profiles_expr
  }
} else {
  profiles_de <- profiles_expr
}
```

```{r compare}
result <- comparePairSets(profiles_expr, profiles_de,
  min_overlap = 2L, verbose = FALSE)
names(result)
```

Inspect a specific section:

```{r compare-detail}
result$event_count[c("description", "test", "p_value", "cliffs_d")]
```

Skipped sections indicate when optional inputs were not provided:

```{r compare-skipped}
result$topology
```

# Meta-Analysis

```{r meta, eval=eval_meta}
# Pool results across multiple comparisons using random-effects meta-analysis
comp1 <- comparePairSets(profiles_expr, profiles_de,
  min_overlap = 2L, verbose = FALSE)
comp2 <- comparePairSets(profiles_expr, profiles_de,
  min_overlap = 2L, verbose = FALSE)
meta <- metaAnalyzePairSets(list(set1 = comp1, set2 = comp2), verbose = FALSE)
names(meta)
```

The `metaAnalyzePairSets()` function requires the **metafor** package and at
least 2 comparisons. Each test section is meta-analyzed independently using
random-effects models.

# Session Info

```{r session-info}
sessionInfo()
```
