---
title: "NMD Mechanism Attribution with Isopair"
author: "Pete Castaldi"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{NMD Mechanism Attribution with Isopair}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
eval_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

# Introduction

Nonsense-mediated mRNA decay (NMD) is triggered when a ribosome encounters a
premature termination codon (PTC) more than 50 nucleotides upstream of an
exon-exon junction. In long-read RNA-seq data, NMD-sensitive isoforms are
identified empirically (e.g., by response to SMG1 inhibition), but the
structural mechanism --- *which splice event created the PTC?* --- requires
computational attribution.

**Isopair** provides a complete workflow for this attribution:

1. **Detect** splice events between isoform pairs (`buildProfiles()`)
2. **Identify** PTCs from exon structure + CDS annotation (`computePtcStatus()`)
3. **Trace** the reading frame through splice events (`analyzeFrameWalk()`)
4. **Attribute** each PTC to a specific splice event and mechanism
   (`attributePtcEvents()` for frameshift/in-frame stop;
   `attribute3UtrSplice()` for same-stop 3'UTR mechanism)

When transcript sequences are available, a fifth step uses
`traceReferenceAtg()` to detect hidden PTCs in isoforms where CDS prediction
software selected a different (non-PTC) reading frame.

This vignette demonstrates the workflow using GENCODE-annotated isoforms from
5 genes, including 3 with PTC-positive isoforms.

# Setup

```{r library, message=FALSE}
library(Isopair)
```

```{r ggplot-setup, eval=eval_ggplot}
library(ggplot2)
theme_set(theme_bw(base_size = 14))
```

# Load and Prepare Data

Isopair ships with example data from GENCODE v49:

```{r load-data}
gtf_path <- system.file("extdata", "example_small.gtf", package = "Isopair")
structures <- parseIsoformStructures(gtf_path, verbose = FALSE)
cds <- extractCdsAnnotations(gtf_path, structures$isoform_id, verbose = FALSE)

cat("Isoforms:", nrow(structures), "from",
    length(unique(structures$gene_id)), "genes\n")
cat("Coding:", sum(cds$coding_status == "coding"), "\n")
```

# Step 1: PTC Detection

`computePtcStatus()` applies the 50-nucleotide rule: a stop codon is premature
if it is >50 nt upstream of the last exon-exon junction in mRNA space.

```{r ptc-detection}
ptc <- computePtcStatus(structures, cds)
cat("PTC status:\n")
print(table(ptc$has_ptc))
```

```{r ptc-table}
# Show PTC+ isoforms with their features
ptc_pos <- ptc[ptc$has_ptc == TRUE, ]
ptc_display <- merge(ptc_pos,
  structures[, c("isoform_id", "gene_id", "strand", "n_exons")])
ptc_display$gene_short <- sub("\\.[0-9]+$", "", ptc_display$gene_id)
ptc_display <- ptc_display[, c("gene_short", "isoform_id", "strand", "n_exons",
                                "orf_length", "ptc_distance", "n_downstream_ejcs")]
knitr::kable(ptc_display,
             caption = "PTC-positive isoforms in the example dataset")
```

`r nrow(ptc_pos)` isoforms have PTCs, with distances ranging from
`r min(ptc_pos$ptc_distance)` to `r max(ptc_pos$ptc_distance)` nt upstream of
the last junction, and 1 to `r max(ptc_pos$n_downstream_ejcs)` downstream
exon-exon junctions.

# Step 2: Build Profiles and Detect Events

We create isoform pairs and detect splice events. For this demonstration, we
select the isoform with the most exons per gene as the reference (a proxy for
the dominant isoform; in a real analysis, use `generatePairsExpression()` or
`generatePairsDE()` which select references by expression level).

```{r build-profiles}
coding_structs <- structures[structures$isoform_id %in%
  cds$isoform_id[cds$coding_status == "coding"], ]

# Build pairs: reference = most exons per gene, comparators = all others
pairs_list <- list()
for (g in unique(coding_structs$gene_id)) {
  g_iso <- coding_structs[coding_structs$gene_id == g, ]
  if (nrow(g_iso) < 2) next
  ref_id <- g_iso$isoform_id[which.max(g_iso$n_exons)]
  comp_ids <- g_iso$isoform_id[g_iso$isoform_id != ref_id]
  for (cid in comp_ids) {
    pairs_list[[length(pairs_list) + 1]] <- data.frame(
      gene_id = g, reference_isoform_id = ref_id,
      comparator_isoform_id = cid, stringsAsFactors = FALSE)
  }
}
pairs <- do.call(rbind, pairs_list)
cat("Pairs:", nrow(pairs), "across",
    length(unique(pairs$gene_id)), "genes\n")

# Build union exons (required by buildProfiles)
ue <- buildUnionExons(structures)

# Build profiles with full event detection
profiles <- buildProfiles(pairs, structures,
                           ue$union_exons, ue$isoform_union_mapping,
                           verbose = FALSE)
cat("Profiles built:", nrow(profiles), "\n")
```

## Extracting Events

`extractAllEvents()` flattens the `detailed_events` list-column from profiles
into a single data frame for analysis:

```{r extract-events}
all_events <- extractAllEvents(profiles)
cat("Total events:", nrow(all_events), "\n")
cat("Event types:\n")
print(sort(table(all_events$event_type), decreasing = TRUE))
```

# Step 3: Reading Frame Analysis

`analyzeFrameWalk()` traces the reading frame through each CDS-overlapping
splice event, tracking the cumulative frame offset. An event that shifts the
frame (bp change not divisible by 3) is flagged as `is_frameshift`. Events
that restore a previous shift are flagged as `is_compensatory`.

```{r frame-walk}
fw <- analyzeFrameWalk(profiles, cds)

cat("Frame walk results:\n")
cat("  Events analyzed:", nrow(fw$events), "\n")
cat("  Pairs with frameshifts:",
    sum(fw$profile_summary$n_frameshifts > 0), "\n")
cat("  Compensatory events:",
    sum(fw$events$is_compensatory, na.rm = TRUE), "\n")
```

```{r frame-walk-detail}
if (nrow(fw$events) > 0) {
  evt_display <- fw$events[, c("gene_id", "comparator_isoform_id",
    "event_type", "direction", "cds_overlap_bp", "signed_cds_change",
    "cumulative_frame_offset", "is_frameshift")]
  evt_display$gene_short <- sub("\\.[0-9]+$", "", evt_display$gene_id)
  knitr::kable(evt_display[, c("gene_short", "comparator_isoform_id",
    "event_type", "direction", "cds_overlap_bp", "signed_cds_change",
    "cumulative_frame_offset", "is_frameshift")],
    caption = "Frame walk events: each row is a CDS-overlapping splice event")
}
```

The `cumulative_frame_offset` column tracks the running reading frame state
(0 = in-frame, 1 or 2 = shifted). When `is_frameshift` is TRUE, the event
(or its splice group) has shifted the frame, which typically creates a
downstream PTC.

# Step 4: PTC Attribution

For each PTC+ pair, we identify which specific splice event caused the
premature stop codon. The attribution uses a three-mechanism framework:

1. **Frameshift**: a splice event shifts the reading frame, creating a stop
   codon in the new frame
2. **In-frame stop**: a splice event introduces exonic sequence containing an
   in-frame stop codon
3. **3'UTR splice**: the reference and comparator share the same stop codon,
   but a downstream splice event repositions an EJC, making the shared stop
   premature

Mechanisms 1 and 2 are handled by `attributePtcEvents()`. Mechanism 3
(same-stop pairs) is handled by `attribute3UtrSplice()`.

## Classify PTC+ Pairs

We first need to determine which PTC+ pairs have a different stop codon
(diff-stop, mechanisms 1-2) versus the same stop codon (same-stop,
mechanism 3). `compareIsoformFrames()` provides the frame category, which
tells us whether the pair shares the same CDS start and whether the reading
frame is preserved.

```{r classify-ptc-pairs}
# Identify PTC+ pairs
ptc_lookup <- setNames(ptc$has_ptc, ptc$isoform_id)
profiles$comp_has_ptc <- ptc_lookup[profiles$comparator_isoform_id]
ptc_plus_pairs <- profiles[profiles$comp_has_ptc == TRUE, ]

cat("PTC+ pairs:", nrow(ptc_plus_pairs), "\n")
```

```{r same-vs-diff-stop}
if (nrow(ptc_plus_pairs) > 0) {
  # Strand-aware stop positions
  stop_pos <- getStrandAwareStop(ptc_plus_pairs$comparator_isoform_id, cds)
  ref_stop <- getStrandAwareStop(ptc_plus_pairs$reference_isoform_id, cds)
  strand_vec <- setNames(
    cds$strand[match(ptc_plus_pairs$comparator_isoform_id, cds$isoform_id)],
    ptc_plus_pairs$comparator_isoform_id)

  same_stop <- stop_pos == ref_stop
  cat("Same-stop pairs:", sum(same_stop, na.rm = TRUE), "\n")
  cat("Diff-stop pairs:", sum(!same_stop, na.rm = TRUE), "\n")
}
```

## Attribute Diff-Stop Pairs (Frameshift / In-Frame Stop)

For diff-stop pairs, `attributePtcEvents()` checks whether the PTC was
caused by a reading frame shift or by an in-frame stop codon introduced by
a splice event. It uses the frame walk results from Step 3 to identify
frameshift events, and coordinate containment to find events spanning
in-frame stops.

```{r attribute-diff-stop}
if (nrow(ptc_plus_pairs) > 0 && any(!same_stop, na.rm = TRUE)) {
  diffstop <- ptc_plus_pairs[!same_stop, ]

  # Compare reading frames to determine frameshift category
  cds_exons <- extractCdsExons(gtf_path, verbose = FALSE)
  fc <- compareIsoformFrames(profiles, cds_exons, cds)
  fs_cats <- c("same_start_frameshift",
               "diff_start_diff_frame_with_frameshift")
  is_fs <- setNames(fc$pair_summary$frame_category %in% fs_cats,
                      fc$pair_summary$comparator_isoform_id)

  diffstop_attr <- attributePtcEvents(
    pairs = diffstop[, c("comparator_isoform_id", "reference_isoform_id")],
    fw_events = fw$events,
    profiles = profiles,
    ptc_genomic_pos = stop_pos[diffstop$comparator_isoform_id],
    strand_vec = strand_vec[diffstop$comparator_isoform_id],
    is_frameshift_vec = is_fs[diffstop$comparator_isoform_id])

  cat("Diff-stop attribution:\n")
  print(table(diffstop_attr$mechanism))
  knitr::kable(diffstop_attr,
    caption = "PTC attribution for diff-stop pairs (mechanisms 1-2)")
}
```

## Attribute Same-Stop Pairs (3'UTR Splice)

For same-stop pairs, the PTC arises not from a different stop codon but from
a splice event downstream of the stop that repositions an exon-exon junction.
`attribute3UtrSplice()` finds the nearest downstream event:

```{r attribute-same-stop}
if (nrow(ptc_plus_pairs) > 0 && any(same_stop, na.rm = TRUE)) {
  samestop <- ptc_plus_pairs[same_stop, ]

  samestop_attr <- attribute3UtrSplice(
    pairs = samestop[, c("comparator_isoform_id", "reference_isoform_id")],
    profiles = profiles,
    stop_genomic_pos = stop_pos[samestop$comparator_isoform_id],
    strand_vec = strand_vec[samestop$comparator_isoform_id])

  cat("Same-stop (3'UTR splice) attribution:\n")
  print(table(samestop_attr$ptc_causing_event))
  knitr::kable(samestop_attr,
    caption = "PTC attribution for same-stop pairs (mechanism 3)")
} else {
  cat("No same-stop PTC+ pairs in this example dataset.\n")
}
```

# Coordinate Utilities

Isopair provides coordinate conversion between genomic and transcript
(mRNA) space. Both functions use 1-based inclusive coordinates (matching
GTF convention):

```{r coordinate-demo}
# Plus-strand gene with 3 exons: [1000,1100], [2000,2200], [3000,3300]
exon_starts <- c(1000L, 2000L, 3000L)
exon_ends <- c(1100L, 2200L, 3300L)

# Genomic position 2050 maps to transcript position:
tx_pos <- genomicToTranscript(2050, exon_starts, exon_ends, "+")
cat("Plus strand: genomic 2050 -> transcript position", tx_pos, "\n")

# Round-trip back to genomic:
back <- transcriptToGenomic(tx_pos, exon_starts, exon_ends, "+")
cat("Round-trip: transcript", tx_pos, "-> genomic", back, "\n")

# Minus strand: transcript reads from high to low genomic coordinates
tx_minus <- genomicToTranscript(2050, exon_starts, exon_ends, "-")
cat("Minus strand: genomic 2050 -> transcript position", tx_minus, "\n")

# Intronic positions return NA:
cat("Intronic position:", genomicToTranscript(1500, exon_starts, exon_ends, "+"), "\n")
```

# UTR Length Analysis

`computeUtrLengths()` computes strand-aware exonic 5'UTR, 3'UTR, and CDS
lengths from structure and CDS annotation:

```{r utr-lengths}
utrs <- computeUtrLengths(structures, cds)
cat("UTR lengths for", nrow(utrs), "coding isoforms:\n")
cat("  Median 5'UTR:", median(utrs$utr5_bp), "bp\n")
cat("  Median 3'UTR:", median(utrs$utr3_bp), "bp\n")
cat("  Median CDS:", median(utrs$cds_bp), "bp\n")
```

```{r utr-ptc-comparison, eval=eval_ggplot, fig.height=5}
utrs$has_ptc <- ptc$has_ptc[match(utrs$isoform_id, ptc$isoform_id)]
utrs_coding <- utrs[!is.na(utrs$has_ptc), ]

utr_long <- rbind(
  data.frame(region = "5'UTR", length = utrs_coding$utr5_bp,
             ptc = ifelse(utrs_coding$has_ptc, "PTC+", "PTC-")),
  data.frame(region = "3'UTR", length = utrs_coding$utr3_bp,
             ptc = ifelse(utrs_coding$has_ptc, "PTC+", "PTC-")))

ggplot(utr_long, aes(x = ptc, y = length, fill = ptc)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1.5) +
  facet_wrap(~region, scales = "free_y") +
  scale_fill_manual(values = c("PTC+" = "#d6604d", "PTC-" = "#4393c3")) +
  labs(title = "UTR Lengths: PTC+ vs PTC- Isoforms",
       subtitle = sprintf("PTC+: %d | PTC-: %d coding isoforms",
                          sum(utrs_coding$has_ptc), sum(!utrs_coding$has_ptc)),
       x = NULL, y = "Exonic length (bp)") +
  theme(legend.position = "none", text = element_text(size = 14))
```

PTC+ isoforms tend to have longer measured 3'UTRs because their premature stop
is upstream of the annotated stop --- the region between the PTC and the
transcript end is classified as "3'UTR" even though part of it was originally
CDS. This artifact is important to account for when interpreting 3'UTR length
in NMD studies.

# Reference ATG Tracing (Hidden PTC Detection)

CDS prediction tools (e.g., TransDecoder2) select the longest high-scoring
ORF, which may not be the ORF the ribosome actually translates. When a splice
event creates a frameshift that truncates the reading frame, the CDS predictor
may select a different ORF whose stop codon has no downstream EJCs --- making
the isoform appear PTC-negative.

`traceReferenceAtg()` addresses this by tracing the reference isoform's CDS
start ATG through each comparator, walking the reading frame to find the first
stop codon, and counting downstream junctions. This requires transcript
sequences:

```{r trace-reference-atg}
# For demonstration, construct minimal sequences for one gene.
# In practice, sequences come from the SQANTI-corrected FASTA.
# We use gene ENSG00000112081.19 (SRSF3), which has PTC+ isoforms.

demo_gene <- "ENSG00000112081.19"
demo_structs <- structures[structures$gene_id == demo_gene, ]
demo_cds <- cds[cds$isoform_id %in% demo_structs$isoform_id, ]
demo_coding <- demo_cds[demo_cds$coding_status == "coding", ]

cat("Gene:", demo_gene, "\n")
cat("Coding isoforms:", nrow(demo_coding), "\n")
for (i in seq_len(nrow(demo_coding))) {
  iso <- demo_coding$isoform_id[i]
  has_p <- if (iso %in% ptc$isoform_id) ptc$has_ptc[ptc$isoform_id == iso] else NA
  cat(sprintf("  %s: CDS %d-%d, PTC=%s\n",
              iso, demo_coding$cds_start[i], demo_coding$cds_stop[i], has_p))
}

# Build synthetic sequences matching the exon structure.
# Each isoform's transcript sequence = concatenated exon sequences.
# We place ATG at the CDS start position and a stop codon at the CDS stop.
build_demo_seq <- function(iso_id, structs, cds_row) {
  str_row <- structs[structs$isoform_id == iso_id, ]
  exon_s <- str_row$exon_starts[[1]]
  exon_e <- str_row$exon_ends[[1]]
  tx_len <- sum(exon_e - exon_s + 1L)

  # Start with random non-stop codons
  seq_chars <- rep("A", tx_len)

  # Map CDS start to transcript position
  cds5 <- if (cds_row$strand == "+") cds_row$cds_start else cds_row$cds_stop
  cds3 <- if (cds_row$strand == "+") cds_row$cds_stop else cds_row$cds_start
  atg_tx <- genomicToTranscript(cds5, exon_s, exon_e, cds_row$strand)
  stop_tx <- genomicToTranscript(cds3, exon_s, exon_e, cds_row$strand)

  if (!is.na(atg_tx) && atg_tx + 2 <= tx_len) {
    seq_chars[atg_tx:(atg_tx + 2)] <- c("A", "T", "G")
  }
  if (!is.na(stop_tx) && stop_tx + 2 <= tx_len) {
    seq_chars[stop_tx:(stop_tx + 2)] <- c("T", "A", "A")
  }

  # Fill CDS region with non-stop codons (GCA repeated)
  if (!is.na(atg_tx) && !is.na(stop_tx) && atg_tx < stop_tx) {
    cds_start <- atg_tx + 3L
    cds_end <- stop_tx - 1L
    if (cds_end >= cds_start) {
      cds_len <- cds_end - cds_start + 1L
      fill <- rep(c("G", "C", "A"), length.out = cds_len)
      seq_chars[cds_start:cds_end] <- fill
    }
  }

  paste(seq_chars, collapse = "")
}

demo_seqs <- setNames(
  sapply(demo_coding$isoform_id, function(iso) {
    build_demo_seq(iso, demo_structs, demo_coding[demo_coding$isoform_id == iso, ])
  }),
  demo_coding$isoform_id)

cat("\nSequences built:", length(demo_seqs), "isoforms\n")
cat("Lengths:", paste(nchar(demo_seqs), collapse = ", "), "\n")
```

```{r run-trace}
# Create pairs: reference (non-PTC) vs comparator (PTC+)
ref_iso <- demo_coding$isoform_id[
  !demo_coding$isoform_id %in% ptc_pos$isoform_id][1]
comp_isos <- demo_coding$isoform_id[
  demo_coding$isoform_id %in% ptc_pos$isoform_id]

if (length(ref_iso) > 0 && length(comp_isos) > 0) {
  trace_pairs <- data.frame(
    reference_isoform_id = ref_iso,
    comparator_isoform_id = comp_isos,
    stringsAsFactors = FALSE)

  trace_result <- traceReferenceAtg(
    pairs = trace_pairs,
    structures = demo_structs,
    cds_metadata = demo_cds,
    sequences = demo_seqs)

  knitr::kable(
    trace_result[, c("reference_isoform_id", "comparator_isoform_id",
                      "ref_atg_exonic_in_comp", "ref_orf_length",
                      "comp_orf_length", "n_downstream_ejc", "category")],
    caption = "Reference-ATG tracing results")

  cat("\nCategories:\n")
  print(table(trace_result$category))
}
```

The `category` column classifies each pair:

- **effectively_ptc**: the reference ATG reading frame terminates early with
  downstream EJCs --- a hidden PTC that CDS prediction missed
- **truncated_no_ejc**: the ORF is shorter but the stop has no downstream EJCs
- **ref_atg_lost**: the reference ATG is not exonic in the comparator
- **no_downstream_ejc**: the ORF is intact or longer, no downstream EJCs

# Summary

The Isopair NMD attribution workflow provides:

| Function | Purpose |
|----------|---------|
| `computePtcStatus()` | PTC detection (50-nt rule) |
| `buildProfiles()` | Splice event detection with genomic coordinates |
| `analyzeFrameWalk()` | Reading frame tracing with frameshift identification |
| `attributePtcEvents()` | Frameshift and in-frame stop attribution |
| `attribute3UtrSplice()` | Same-stop (3'UTR splice) attribution |
| `traceReferenceAtg()` | Hidden PTC detection via reference reading frame |
| `genomicToTranscript()` / `transcriptToGenomic()` | Coordinate conversion |
| `getStrandAwareStop()` | Strand-correct stop codon position |
| `computeUtrLengths()` | Exonic UTR and CDS lengths |
| `extractAllEvents()` | Flatten event list-columns for analysis |

This pipeline enables systematic NMD mechanism characterization at scale,
connecting individual splice events to their downstream consequences for
translation and mRNA stability.

# Session Info

```{r session-info}
sessionInfo()
```
